Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-05-12 paper.tex
Summarise mutations

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
For validation purposes, we compare the genetic
relationships in our ARGs with phylogenetic relationships
inferred by the NextStrain project \citep{Hadfield2018-ef}.
To enable this comparison, we ``simplified''~\citep{Kelleher2018-xc}
the sc2ts ARGs to a ``backbone'' containing only samples present
in a specific NextStrain tree. For this purpose, we picked an
``all-data''  NextStrain tree restricted to the time period covered
by the appropriate ARG. This results in a relatively small
set of shared samples, almost entirely restricted to non-recombinant lineages.
These simplified backbone ARGs exhibit very
little recombination (6 and 5 trees across the genome for the Wide and Long
ARGs, respectively).
% JK: This leaves a big question hanging - why do we ahave recombination
% in here at all? It's probably just artefactual, right?

@@

\subsection{Mutational spectrum}
\label{sec:mutation_spectrum}
 In addition to recombination events, the ARGs record mutation events along the
SARS-CoV-2 phylogeny. \cite{Yi2021-sc} reconstructed a SARS-CoV-2 phylogeny of
over 350,000 genomes sampled globally from December 24, 2019, to January 12,
2021, and systematically catalogued all the mutations occurring along the
phylogeny. Notably, they described a mutational spectrum in which C-to-U and
G-to-U mutations occur more frequently than U-to-C and U-to-G mutations,
respectively.

@@

\begin{tabular}[b]{llrlr}\toprule
            & \multicolumn{2}{l}{Wide ARG} & \multicolumn{2}{l}{Long ARG} \\
% \midrule
    \cmidrule{2-5}
Total      & \multicolumn{2}{l}{1,213,193} & \multicolumn{2}{l}{1,062,072} \\
Transitions & 873,487   & (72.00\%) & 783,773  & (73.80\%) \\
Transversions & 326,053 & (26.88\%) & 270,333  & (25.45\%) \\
Insertions  & 6191  & (0.51\%) & 2814   & (0.26\%) \\
Deletions   & 7462  & (0.62\%) & 5152   & (0.49\%) \\
Recurrent   & 74,719 & (6.16\%) & 50,099 & (4.72\%) \\
Reversions  & 72,617 & (5.99\%) & 48,226 & (4.54\%) \\
Private     & 758,903 & (62.55\%) &  767,111 & (72.23\%)\\
\bottomrule
\end{tabular}
&
\includegraphics[width=0.35\textwidth]{figures/mutational_spectra.pdf}\\
(A) & (B)\\
\end{tabularx}
\caption{\label{fig:mutational_spectra}
(A) Summary of mutations in the Long and Wide ARGs.
(B) Mutational spectrum of SARS-CoV-2 in
the wide ARG compared to \cite{Yi2021-sc}. Mutations are
categorised by type (i.e., ancestral state $>$ derived state). The percentages
of each mutation type from the wide ARG are represented by blue bars and the
percentages from Yi et al. by orange bars, with the darker colours representing
one direction (e.g., C$>$U) and the lighter colours the reverse (e.g., U$>$C).}
\end{figure}

For validation purposes, we compare the mutational spectrum from the wide ARG
with the mutational spectrum of \citet{Yi2021-sc}. We categorised all of the
single nucleotide mutations by type (defined by the ancestral state and derived
state), excluded mutations inherited by only a single sample (which are
enriched for sequencing errors and artefacts), and tallied them up by type.
Similarly, we took the data for single nucleotide mutations from
\url{https://github.com/ju-lab/SC2_evol_signature} \citep{Yi2021-sc}, excluded
mutations occurring along terminal branches, and tallied them up by type.
Figure~\ref{fig:mutational_spectra} shows that the mutational spectrum from the
wide ARG (based on 448,825 mutations) matches that reported by \citet[based on
92,344 mutations]{Yi2021-sc}. In both spectra, C-to-U mutations and G-to-U
mutations occur more frequently than U-to-C and U-to-G, respectively. Similar
results are obtained when including the mutations inherited by only a single
sample or those occurring on terminal branches (data not shown).

% If we do this, goes here
% \subsection{Mutations under selection}
% XX

>>>>

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
%% The ARG is mostly tree-like
Compared to organisms like humans which recombine every generation,
in SARS-CoV2 recombination is rare, with recombination nodes accounting
for <0.3\% of our inferred ancestry (Table~\ref{tab:args}). As a result,
relationships between strains can often be represented by a
single phylogenetic tree, particularly when looking at
a subset of sampled genomes e.g. restricted by time or limited to close
relatives. We expect the ARG to be particularly tree-like
early in the pandemic, when co-infection was less likely.

A classic tree-based summary of SARS-CoV2 ancestry is provided
by the NextStrain project \citep{Hadfield2018-ef}. The trees
available from Nextstrain are based on small subsamples of the
dataset, and early in the pandemic tend to be restricted to
a sample of strains which are not thought to be recombinants.
For validation purposes, we our compare our ARGs with
a downloaded ``all-data'' NextStrain tree, restricted to the time period
covered by each ARG. To enable this, we ``simplified''~\citep{Kelleher2018-xc}
the sc2ts ARGs to a ``backbone'' containing only samples present
the NextStrain tree. This results in a small set of shared
samples (180 for the wide ARG, 88 for the long ARG), none of which are
designated as Pango recombinant lineages by Nextclade (see subsection
\ref{sec:pango_x_lineages}).

These sc2ts backbone phylogenies do contain small amounts of recombination

@@

\subsection{Mutational spectrum}
\label{sec:mutation_spectrum}
The ARGs inferred by \texttt{sc2ts} and represented using the \texttt{tskit}
library (Section~\ref{sec:args}) are a joint estimate of the processes
of inheritance, recombination and mutation. Unlike most approaches to
phylogenetic analysis, mutations are included in the \texttt{tskit}
data model alongside
the topological representation of genetic inheritance.
This has many
advantages, for example allowing us to compute statistics of the observed
sequences efficiently~\citep{Kelleher2016-wk,Ralph2020-efficiently} and
to provide high-levels of data compression~\citep{Kelleher2019-ba}.
The same idea has recently been used to represent
SARS-CoV-2 data in UShER's ``mutation annotated tree'' format~\citep{Turakhia2022-it}.

@@

\begin{tabular}[b]{llrlr}\toprule
            & \multicolumn{2}{l}{Wide ARG} & \multicolumn{2}{l}{Long ARG} \\
% \midrule
    \cmidrule{2-5}
Total      & \multicolumn{2}{l}{1,213,193} & \multicolumn{2}{l}{1,062,072} \\
Private     & 758,903 & (62.55\%) &  767,111 & (72.23\%)\\
    \cmidrule{2-5}
Transitions & 873,487   & (72.00\%) & 783,773  & (73.80\%) \\
Transversions & 326,053 & (26.88\%) & 270,333  & (25.45\%) \\
    \cmidrule{2-5}
Insertions  & 6191  & (0.51\%) & 2814   & (0.26\%) \\
Deletions   & 7462  & (0.62\%) & 5152   & (0.49\%) \\
    \cmidrule{2-5}
Recurrent   & 74,719 & (6.16\%) & 50,099 & (4.72\%) \\
Reversions  & 72,617 & (5.99\%) & 48,226 & (4.54\%) \\
\bottomrule
\end{tabular}
&
\includegraphics[width=0.35\textwidth]{figures/mutational_spectra.pdf}\\
(A) & (B)\\
\end{tabularx}
\caption{\label{fig:mutational_spectra}
(A) Summary of mutations in the Long and Wide ARGs. Private mutations
occur on terminal branches. Insertions are mutations in which the
inherited state is the gap state ``-'' and the derived state is a
nucleotide, and vice versa for deletions. Recurrent mutations are
those that inherit from at least one other mutation (at that site),
and a reversion is a recurrent mutation that reverses the state change
of the mutation immediately ancestral to it.
(B) Mutational spectrum in
the Wide ARG compared to \cite{Yi2021-sc}. Mutations are
categorised by type (i.e., inherited state $>$ derived state). The percentages
of each mutation type from the wide ARG are represented by blue bars and the
percentages from Yi et al. by orange bars, with the darker colours representing
one direction (e.g., C$>$U) and the lighter colours the reverse (e.g., U$>$C).}
\end{figure}

The properties of the mutations inferred in the Wide and Long
ARGs are summarised in Figure~\ref{fig:mutational_spectra}A. In both
cases we have a large number of mutations, and a majority of these
(Wide ARG: 62.55\%, Long ARG: 72.23) are private to a single
sample, i.e.\ on terminal branches.
Although the average number
of mutations per sample is small
(Wide ARG: 0.77, Long ARG: 1.38; c.f.\ Table~\ref{tab:args}), the average number per site
is large (Wide ARG: 41.23, Long ARG: 36.10; c.f.\ Table~\ref{tab:args}). This is
partly driven by some highly mutagenic sites (e.g., site 28271 has over 7,000
mutations in the Wide ARG, c.f.\ Figure~\ref{fig:breakpoint-distribution}), which
may be artefactual.
The main weakness of the current version of \texttt{sc2ts} is an excess
of inferred reversion mutations, with 5.99\% of all mutations in the Wide
ARG (Long ARG: 4.54\%) reverting the state change of the immediately
ancestral mutation. The majority of these mutations are likely to be
artefactual, and a key goal of future development is to improve
our parsimony heuristics (Section~\ref{sec:parsimony-heuristics}) to
address this issue.

Despite the presence of some artefactual mutations,
the properties of the mutations inferred by \texttt{sc2ts} largely follow
established results.
In Figure~\ref{fig:mutational_spectra}B we compare the mutational spectrum
in the Wide ARG to the results of \cite{Yi2021-sc},
who reconstructed a SARS-CoV-2 phylogeny of
over 350,000 genomes sampled globally from December 24, 2019, to January 12,
2021, and classified the mutations occurring along the phylogeny.
We categorised all of single nucleotide
mutations in the Wide ARG by type (defined by the inherited and derived states),
excluded mutations inherited by only a single sample (which are
enriched for sequencing errors and artefacts).
Similarly, we took the data for single nucleotide mutations from
\citep[][\url{https://github.com/ju-lab/SC2_evol_signature}]{Yi2021-sc}, excluded
mutations occurring along terminal branches, and tallied them up by type.
Figure~\ref{fig:mutational_spectra}B shows that the mutational spectrum from the
wide ARG (based on 448,825 mutations) matches that reported by \citet[based on
92,344 mutations]{Yi2021-sc}. In both spectra, C-to-U mutations and G-to-U
mutations occur more frequently than U-to-C and U-to-G, respectively. Similar
results are obtained when including the mutations inherited by only a single
sample or those occurring on terminal branches (data not shown).
